A comparison of classification algorithms applied to the Adult Database

### 2009-12-10 report/final.tex
minor fixes in the report.

Deducing from the previous section, this is a classification problem. Given an adult
with the mentioned attributes we want to classify him into one of the following groups:
one gains more than 50000 dollars per year, the other group does not.

@@

For the second step in building a bayesian network classifier, we used a
simple estimator to estimate the conditional probability distributions.
This simple estimator produces
direct estimates of the conditional probabilities. \cite{bayes_weka}

@@

@@

We experimented the algorithm using various combinations of those two parameters.
Tables \ref{tbl:results_adtree_discr} and \ref{tbl:results_adtree_raw} show
results for both discretized data and raw data, respectively.
An example tree built using non-discretized data and with the parameters \textbf{-E -2 -B 10}
is displayed in Listing \ref{decision_tree_adtree}.

@@

  \end{tabular}
  \caption{Results for ADTree for discretized data.}
  \label{tbl:results_adtree_discr}
  \end{center}
\end{table}

\begin{table}[ht]
  \begin{center}
  \begin{tabular}{ | l | c | c | c | c | c |}
    \hline
    \textbf{Parameters} & \textbf{Train error rate} & \textbf{Test error rate} & \textbf{Time} & \textbf{Nodes} & \textbf{Leaves} \\ \hline
    -E -3 -B 2 & 19.90 & 19.69 & 0m4.166s & 7 & 5 \\ \hline
    -E -3 -B 5 & 18.43 & 18.58 & 0m7.070s & 16 & 11 \\ \hline
    -E -3 -B 10 & 14.98 & 19.68 & 0m14.188s & 31 & 21 \\ \hline
    -E -3 -B 15 & 14.78 & 20.18 & 0m24.777s & 46 & 31 \\ \hline
    -E -3 -B 20 & 14.48 & 19.88 & 0m36.194s & 61 & 41 \\ \hline
    -E -3 -B 25 & 14.08 & 18.87 & 0m51.563s & 76 & 51 \\ \hline
    -E -3 -B 30 & 13.80 & 18.49 & 1m11.919s & 91 & 61 \\ \hline
    -E -3 -B 35 & 13.74 & 18.65 & 1m32.886s & 106 & 71 \\ \hline
    -E -3 -B 40 & 13.65 & 18.98 & 1m54.094s & 121 & 81 \\ \hline

    -E -2 -B 2 & 19.90 & 19.69 & 0m3.995s & 7 & 5 \\ \hline
    -E -2 -B 5 & 18.43 & 18.58 & 0m5.502s & 16 & 11 \\ \hline
    -E -2 -B 10 & 15.14 & 20.07 & 0m8.662s & 31 & 21 \\ \hline
    -E -2 -B 15 & 14.67 & 20.26 & 0m11.982s & 46 & 31 \\ \hline
    -E -2 -B 20 & 14.67 & 19.95 & 0m15.543s & 61 & 41 \\ \hline
    -E -2 -B 25 & 14.43 & 19.88 & 0m19.500s & 76 & 51 \\ \hline
    -E -2 -B 30 & 14.34 & 19.85 & 0m22.929s & 91 & 61 \\ \hline
    -E -2 -B 35 & 14.26 & 20.00 & 0m27.735s & 106 & 71 \\ \hline
    -E -2 -B 40 & 14.05 & 19.71 & 0m32.832s & 121 & 81 \\ \hline

    -E -1 -B 2 & 19.90 & 19.69 & 0m4.093s & 7 & 5 \\ \hline
    -E -1 -B 5 & 18.43 & 18.58 & 0m5.414s & 16 & 11 \\ \hline
    -E -1 -B 10 & 15.14 & 20.07 & 0m8.459s & 31 & 21 \\ \hline
    -E -1 -B 15 & 14.67 & 20.26 & 0m11.937s & 46 & 31 \\ \hline
    -E -1 -B 20 & 14.67 & 19.95 & 0m15.430s & 61 & 41 \\ \hline
    -E -1 -B 25 & 14.43 & 19.88 & 0m19.413s & 76 & 51 \\ \hline
    -E -1 -B 30 & 14.34 & 19.85 & 0m22.973s & 91 & 61 \\ \hline
    -E -1 -B 35 & 14.26 & 20.00 & 0m27.246s & 106 & 71 \\ \hline
    -E -1 -B 40 & 14.05 & 19.72 & 0m32.555s & 121 & 81 \\ \hline

    -E 0 -B 2 & 19.90 & 19.69 & 0m4.039s & 7 & 5 \\ \hline
    -E 0 -B 5 & 18.43 & 18.58 & 0m4.901s & 16 & 11 \\ \hline
    -E 0 -B 10 & 15.15 & 20.10 & 0m6.701s & 31 & 21 \\ \hline
    -E 0 -B 15 & 14.80 & 20.65 & 0m8.409s & 46 & 31 \\ \hline
    -E 0 -B 20 & 14.50 & 20.27 & 0m9.858s & 61 & 41 \\ \hline
    -E 0 -B 25 & 14.37 & 20.41 & 0m11.057s & 73 & 49 \\ \hline
    -E 0 -B 30 & 14.21 & 20.28 & 0m12.712s & 88 & 59 \\ \hline
    -E 0 -B 35 & 13.99 & 19.45 & 0m14.145s & 103 & 69 \\ \hline
    -E 0 -B 40 & 14.04 & 20.21 & 0m15.273s & 118 & 79 \\ \hline

  \end{tabular}
  \caption{Results for ADTree for non-discretized data.}
  \label{tbl:results_adtree_raw}
  \end{center}
\end{table}

From tables \ref{tbl:results_adtree_discr} and \ref{tbl:results_adtree_raw}
we can see that the number of boosting iterations affects
both the execution time (directly proportional) and the error rate.
The best error rate for raw data is using the parameters \textbf{-E -1 -B 5} giving $\approx$18.58\% for the test set.

For nearly all parameters the algorithm gives best results when using numeric attributes instead
of categories. Non-discretized data also generates larger trees.

@@

The information gain and modified entropy approaches to error estimations are
only available for the algorithm producing ordered set of rules. The approach
building an unordered set of rules with a naive error estimation lead to a stack
overflow error in the application. One thing one can note is that the process of
building classification rules takes, in average, longer than the one of building
classification trees. Nevertheless, the results are arguably more expressive and
the error rates are low (the best being 15.2\% for a previously unkown set of 
examples). Also note that the variant of building an ordered set of rules with
an information gain error estimate produces rules faster and with an acceptable
error rate (18.9\%). To analyze the expressiveness of the rules produced, a set
of rules that cover the most examples of the training data are described in 
listing \ref{cn2_rules} (using the ordered algorithm with the laplacian error 
estimation).

@@

\clearpage
\begin{lstlisting}[language=c,frame=single,breaklines=true,basicstyle=\footnotesize\ttfamily,caption={Decision tree built by ADTree with -E -2 -B 10},label=decision_tree_adtree]
: -0.574
|  (1)marital_status = married_civ_spouse: 0.467
|  (1)marital_status != married_civ_spouse: -0.756
|  (2)capital_gain < 7073.5: -0.11
|  |  (5)capital_loss < 1748: -0.047
|  |  (5)capital_loss >= 1748: 0.996
|  |  (7)education = masters: 0.714
|  |  (7)education != masters: -0.055
|  (2)capital_gain >= 7073.5: 2.636
|  (3)age < 27.5: -0.891
|  (3)age >= 27.5: 0.11
|  (4)hours_per_week < 41.5: -0.183
|  (4)hours_per_week >= 41.5: 0.291
|  (6)education = bachelors: 0.418
|  (6)education != bachelors: -0.101
|  (8)education = prof_school: 0.986
|  (8)education != prof_school: -0.023
|  (9)education = doctorate: 1.067
|  (9)education != doctorate: -0.021
|  (10)occupation = exec_managerial: 0.335
|  (10)occupation != exec_managerial: -0.062
\end{lstlisting}

>>>>

The adult database poses a classification problem. Given an adult
with some attributes we want to classify him into one of the following groups:
one gains more than 50000 dollars per year, the other group does not.

@@

@@

For the second step in building a bayesian network classifier, we used a
simple estimator to estimate the conditional probability distributions.
This simple estimator produces
direct estimates of the conditional probabilities \cite{bayes_weka}.

@@

We experimented the algorithm using various combinations of those two parameters.

@@

  \end{tabular}
  \caption{Results for ADTree.}
  \label{tbl:results_adtree}
  \end{center}
\end{table}

From Table \ref{tbl:results_adtree} we can see that the number of boosting iterations affects
both the execution time (directly proportional) and the error rate.
The best error rate is for the parameters \textbf{-E -2 -B 20} giving $\approx$19.80\% for the test set.

@@

The information gain and modified entropy approaches to error estimations are
only available for the algorithm producing ordered set of rules. The approach
building an unordered set of rules with a naive error estimation lead to a stack
overflow error in the application. One thing one can note is that the process of
building classification rules takes, in average, longer than the one of building
classification trees. Nevertheless, the results are arguably more expressive and
the error rates are low (the best being 15.2\% for a previously unkown set of 
examples). Also note that the variant of building an ordered set of rules with
an information gain error estimate produces rules faster and with an acceptable
error rate (18.9\%). Models generating an ordered set of rules with the 
laplacian or naive error estimate overfit the training data. To analyze the 
expressiveness of the rules produced, a set of rules that cover the most 
examples of the training data are described in listing \ref{cn2_rules} (using 
the ordered algorithm with the laplacian error estimation).

@@
